\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-09 papers/jss/article.Rnw
small edits paper does NOT 'compile' without network connectivity which is BAD BAD BAD BAD
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@734 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer system, possibly
spanning different operating systems?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
% DE: need to define serialization?
programming language in use and thus lock the user into a single
environment.  CSV files can be read and written by many applications
and so are often used for exporting tabular data.  However, CSV files
have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and ambiguities in the format involving special
characters.  JSON is another widely-supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Because the schema information
is not kept separately, multiple JSON messages of the same type
needlessly duplicate the field names with each message.
%
%
%
A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
\citep{msgpackR} and BSON \citep{rmongodb} both have R interfaces, but
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time, or when application logic and
requirement changes dictate update to the message format.
% DE: Need to talk about XML ?

@@

<<>>=
a <- new(protobuf_unittest.TestAllTypes)
a$optional_bool <- TRUE
a$optional_bool <- FALSE
<<eval=F>>=
a$optional_bool <- NA
<<echo=FALSE,eval=TRUE,print=TRUE>>=
try(a$optional_bool <- NA,silent=TRUE)
@

@@

<<>>=
as.integer(2^31-1)
<<eval=F>>=
as.integer(2^31 - 1) + as.integer(1)
<<echo=FALSE,eval=TRUE,print=TRUE>>=
try(as.integer(2^31 - 1) + as.integer(1))

@@

# Check that no information was lost
identical(output, MASS::Animals)
@
This code suggests a method for exchanging objects between R servers, however this can 
also be done without protocol buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

@@

<<>>=
#requires httr >= 0.2.99
library(httr)
library(RProtoBuf)

>>>>

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer system, possibly
spanning different operating systems?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
% DE: need to define serialization?
programming language in use and thus lock the user into a single
environment.  CSV files can be read and written by many applications
and so are often used for exporting tabular data.  However, CSV files
have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and ambiguities in the format involving special
characters.  JSON is another widely-supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Because the schema information
is not kept separately, multiple JSON messages of the same type
needlessly duplicate the field names with each message.
Lastly, XML is a well-established and widely-supported protocol with the ability to define
just about any arbitrarily complex schema. However, it pays for this
complexity with comparatively large and verbose messages, and added
complexities at the parsing side. 
%
%
%
A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
and BSON both have R interfaces \citep{msgpackR,rmongodb}, but
% DE Why do we cite these packages, but not the numerous JSON packages?
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time, or when application logic and
requirement changes dictate update to the message format.
% DE: Need to talk about XML -- added a few lines at previous paragraph

@@

<<>>=
a <- new(protobuf_unittest.TestAllTypes)
a$optional_bool <- TRUE
a$optional_bool <- FALSE
<<eval=FALSE>>=
a$optional_bool <- NA
<<echo=FALSE,eval=TRUE,print=TRUE>>=
try(a$optional_bool <- NA,silent=TRUE)
@

@@

<<>>=
as.integer(2^31-1)
<<eval=FALSE>>=
as.integer(2^31 - 1) + as.integer(1)
<<echo=FALSE,eval=TRUE,print=TRUE>>=
try(as.integer(2^31 - 1) + as.integer(1))

@@

# Check that no information was lost
identical(output, MASS::Animals)
@

This code suggests a method for exchanging objects between R servers, however this can 
also be done without protocol buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

@@

<<eval=FALSE>>=
#requires httr >= 0.2.99
library(httr)
library(RProtoBuf)
